Dark Control: The Default Mode Network as a Reinforcement Learning Agent

### 2018-08-07 paper/paper.tex
addressed comments of rev. #2

The present work adopts the perspective of
a human \textit{agent} faced with the choice of the next actions
and guided by outcomes
of really happened, hypothetically imagined, and
expected futures to optimize behavioral performance.
Formally, we propose reinforcement learning to be a particularly attractive framework
to describe, quantify, DMN function and the brain.
An intelligent agent improves the interaction with the environment
by continuously updating its computation of value estimates and action predispositions
through integration of feedback outcomes.
That is, ``[agents], with their actions, modify the environment and in doing so partially determine their next stimuli, in particular stimuli that are necessary for triggering the next action''
\citep{pezzulo2011grounding}.
Agents with other behavioral policies therefore sample different
distributions of action-perception trajectories \citep{ghavamzadeh2015bayesian}.
Henceforth,
\textit{control} refers to the influence that an agent exerts when interacting
with the environment to reach preferred states.

@@

\subsubsection{Accumulated rewards and policies}
The behavior of the agent is governed by a \textit{policy}, which maps states of the world
to probability distributions over candidate actions.
Starting at time $t=0$,
following a policy $\pi$ generates a trajectory of action choices:
\begin{eqnarray*}
  \begin{split}
    \textbf{choose action: }a_0 &\sim \pi(a|s_0)\\
    \textbf{ observe transition: }s_1 &\sim p(s|s_0,a_0)\textbf{ and collect reward }R_0 = r(s_0, a_0, s_1)\\
    \textbf{choose action: }a_1 &\sim \pi(a|s_1)\\
    \textbf{ observe transition: }s_2 &\sim p(s|s_1,a_1), \textbf{ and collect reward }R_1 = r(s_1, a_1,s_2)\\
    \vdots\\
    \textbf{choose action: }a_{t} &\sim \pi(a|s_{t})\\
    \textbf{ observe transition: }s_{t+1} &\sim p(s|s_{t},a_{t}), \textbf{ and collect reward }R_{t} = r(s_{t}, a_{t},s_{t+1})\\
    \vdots
  \end{split}
\end{eqnarray*}
We assume time-invariance
in that we expect the dynamics of the process
to be equivalent over sufficiently long time windows of equal length (i.e., stationarity).
Since an action executed in the present moment might have repercussions in the far future, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s, a)$, but a
\textit{cumulative reward} estimate which takes into account expected reward
from action choices in the future.
A common approach to
modeling this gathered outcome is the time-discounted cumulative reward% function
\begin{equation}
  \label{eq:cumr}
  G^\pi = \sum_{t=0}^{\infty}\gamma^{t}R_t = R_0 + \gamma R_1 + \gamma^2 R_2 + \ldots + \gamma^tR_t + \ldots
\end{equation}
This random variable
  \footnote{Random as it depends both on the environment's dynamics and the
  policy $\pi$ being executed. The exponential delay discounting function used here refers to the usual formulation in the field of reinforcement learning, although psychological experiments may also reveal other discounting regimes \citep{green2004discounting}.}
measures the cumulative reward of
following an action policy $\pi$.
%
Note that value buffering may be realized in the vmPFC.
This DMN region has direct connections to
to the NAc, known to be involved in reward evaluation.

@@

  \subsubsection{Self-training and the loss function}
  \label{sec:self}
Successful learning in brains and computer algorithms may not be possible without a
defined learning goal \textemdash the \textit{loss function}.
The action $a$ chosen in state $s$ according to the policy matrix defined in
\eqref{eq:policy} yields a reward $r$ collected by the agent,
% (measured on their internal currency of worldly values ?),
after which the environment transitions to a new state $s' \in \mathcal S$.
% \textbf{XXX Please try to smoothen the previous sentence}
One such cycle yields a new \textit{experience} $e = (s,a,r,s')$.
Each cycle represents a behavior unit of the agent
and is recorded in replay memory buffer \textemdash
which we hypothesize to be subserved by the
HC \textemdash, possibly discarding the oldest entries to make space:
$\mathcal D \leftarrow \text{append}(\mathcal D, e)$.
At time step $k$, the agent seeks an update $\theta_{k} \leftarrow \theta_{k-1} + \delta \theta_{k}$ of the parameters for its approximate model of the $Q$-value function. This warrants a learning process
and definition of a loss function.
% \subsubsection{Stochastic gradient in the DMN via experience replay}
The Bellman equation \eqref{eq:bellman} provides a way to obtain such a loss function \eqref{eq:oracle} as we outline in the following.
% A practically successful RL approach
% needs ways to efficiently approximate the state-action value function
% to allow for scaling to large state and action spaces.
Experience replay consists in sampling
% (uniform or importance-weighted
% \footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.})
batches of experiences $e$
$(s, a, r, s') \sim \mathcal D$ from the replay memory $\mathcal D$.
The agent then tries to approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by the Bellman equation \eqref{eq:bellman}, namely
\begin{equation}
  y_k := y_k(s,a,s') =  r + \gamma \max_{a'} \tilde{Q}(s', a'|\theta_{k-1}),
\end{equation}
with the estimation of a parametrized regression model $(s,a)
\mapsto \tilde{Q}(s, a|\theta_{k-1})$.
From a neurobiological perspective, experience replay can be manifested as the re-occurrence of neuron spiking sequences that have also been measured during specific prior actions and environmental states. The HC is a strong candidate for contributing to such neural reinstantiation of behavioral episodes as neuroscience experiments have repeatedly indicated in rats, mice, cats, rabbits, songbirds, and monkeys \citep{buhry2011,nokia2010,dave2000,skaggs2007}.

@@

\small
% \bibliographystyle{splncs03}
% \bibliographystyle{natbib}
\bibliographystyle{plainnat}
\bibliography{refs}

>>>>

The present work adopts the perspective of
a human \textit{agent} faced with the choice of the next actions
and guided by outcomes to optimize behavioral performance. These outcomes
can be experienced, hypothetically imagined, or simply future expectations.
Formally, we propose reinforcement learning as a particularly attractive framework
for describing and quantifying the DMN function.
An intelligent agent improves the interaction with the environment
by continuously updating its computation of value estimates and action predispositions
through integration of feedback outcomes.
That is, ``[agents], with their actions, modify the environment and in doing so
partially determine their next stimuli, in particular stimuli that are necessary
for triggering the next action''
\citep{pezzulo2011grounding}.
Agents with other behavioral policies therefore sample different
distributions of action-perception trajectories \citep{ghavamzadeh2015bayesian}.
Henceforth,
\textit{control} refers to the influence that an agent exerts when interacting
with the environment to reach preferred states.

@@

\subsubsection{Accumulated rewards and policies}
The behavior of the agent is governed by a \textit{policy}, which maps states of the world
to probability distributions over candidate actions.
Starting at time $t=0$,
following a policy $\pi$ generates a trajectory of action choices:
\begin{eqnarray*}
  \begin{split}
    \textbf{choose action: }a_0 &\sim \pi(a|s_0)\\
    \textbf{ observe transition: }s_1 &\sim p(s|s_0,a_0)\textbf{ and collect
      reward }R_0 = r(s_0, a_0, s_1)\\
    \textbf{choose action: }a_1 &\sim \pi(a|s_1)\\
    \textbf{ observe transition: }s_2 &\sim p(s|s_1,a_1), \textbf{ and collect
      reward }R_1 = r(s_1, a_1,s_2)\\
    \vdots\\
    \textbf{choose action: }a_{t} &\sim \pi(a|s_{t})\\
    \textbf{ observe transition: }s_{t+1} &\sim p(s|s_{t},a_{t}), \textbf{ and
      collect reward }R_{t} = r(s_{t}, a_{t},s_{t+1})\\
    \vdots
  \end{split}
\end{eqnarray*}
We assume time-invariance
in that we expect the dynamics of the process
to be equivalent over sufficiently long time windows of equal length (i.e., stationarity).
Since an action executed in the present moment might have repercussions in the
far future, it turns out that the quantity to optimize is not the instantaneous
rewards $r(s, a)$, but a \textit{cumulative reward} estimate which takes into
account expected reward from action choices in the future. A common approach to
modeling this gathered outcome is the time-discounted cumulative reward% function
\begin{equation}
  \label{eq:cumr}
  G^\pi = \sum_{t=0}^{\infty}\gamma^{t}R_t = R_0 + \gamma R_1 + \gamma^2 R_2 + \ldots + \gamma^tR_t + \ldots
\end{equation}
This random variable
  \footnote{Random as it depends both on the environment's dynamics and the
  policy $\pi$ being executed. The exponential delay discounting function used
  here refers to the usual formulation in the field of reinforcement learning,
  although psychological experiments may also reveal other discounting regimes
  \citep{green2004discounting}.}
measures the cumulative reward of
following an action policy $\pi$.
%
Note that value buffering may be realized in the vmPFC.
This DMN region has direct connections to the NAc, known to be involved in
reward evaluation~\citep{carlezon2009}.

@@

  \subsubsection{Self-training and the loss function}
  \label{sec:self}
Successful learning in brains and computer algorithms may not be possible without a
defined learning goal \textemdash the \textit{loss function}.
The action $a$ chosen in state $s$ according to the policy matrix defined in
\eqref{eq:policy} yields a reward $r$ collected by the agent,
% (measured on their internal currency of worldly values ?),
after which the environment transitions to a new state $s' \in \mathcal S$.
% \textbf{XXX Please try to smoothen the previous sentence}
One such cycle yields a new \textit{experience} $e = (s,a,r,s')$.
Each cycle represents a behavior unit of the agent
and is recorded in replay memory buffer \textemdash
which we hypothesize to be subserved by the
HC \textemdash, possibly discarding the oldest entries to make space:
$\mathcal D \leftarrow \text{append}(\mathcal D, e)$.
At time step $k$, the agent seeks an update $\theta_{k} \leftarrow \theta_{k-1}
+ \delta \theta_{k}$ of the parameters for its approximate model of the
$Q$-value function. This warrants a learning process and definition of a loss
function. % \subsubsection{Stochastic gradient in the DMN via experience replay}
The Bellman equation \eqref{eq:bellman} provides a way to obtain such a loss
function \eqref{eq:oracle} as we outline in the following.
% A practically successful RL approach
% needs ways to efficiently approximate the state-action value function
% to allow for scaling to large state and action spaces.
Experience replay consists in sampling
% (uniform or importance-weighted
% \footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.})
batches of experiences $e$
$(s, a, r, s') \sim \mathcal D$ from the replay memory $\mathcal D$.
The agent then tries to approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by the
Bellman equation \eqref{eq:bellman}, namely
\begin{equation}
  y_k := y_k(s,a,s') =  r + \gamma \max_{a'} \tilde{Q}(s', a'|\theta_{k-1}),
\end{equation}
with the estimation of a parametrized regression model $(s,a)
\mapsto \tilde{Q}(s, a|\theta_{k-1})$.
From a neurobiological perspective, experience replay can be manifested as the
re-occurrence of neuron spiking sequences that have also been measured during
specific prior actions and environmental states. The HC is a strong candidate
for contributing to such neural reinstantiation of behavioral episodes as
neuroscience experiments have repeatedly indicated in rats, mice, cats, rabbits,
songbirds, and monkeys \citep{buhry2011,nokia2010,dave2000,skaggs2007}. Refer to
Fig. \ref{fig:rl_process_chart}.

@@

\small
% \bibliographystyle{splncs03}
% \bibliographystyle{natbib}
% \bibliographystyle{plainnat}
\bibliographystyle{abbrvnat}
\bibliography{refs}
